Quantum-Classical Phase Transition for Exciton-Polariton Josephson Junctions

### 2017-01-09 writing/phase_transition/phase_transition.tex
Bibliography OK

For last decade investigations of exciton polariton Bose-Einstein condensates (BEC) in various type of semiconductor microstructures represent one of mainstreams of current photonic and semiconductor technologies, see e.g. \cite{1,2}.
The microcavity exciton polaritons are bosonic quasiparticles representing  admixtures of the quantized cavity mode and quantum wells excitons.
Practically, semiconductor microcavities are promising for various optoelectronic applications where quantum  matter-field interface plays crucial role.
In particular, demonstrations of polariton laser with electrical pump \cite{3,4}, amplifiers \cite{5}, switches \cite{6}, transistors \cite{7}, polariton circuits and optical logic elements \cite{8,9}.
Although exciton polariton condensates exhibits Bose-Einstein statistics under the phase transition condition and macroscopic occupation of the ground state at certain pumping rate that is less than for convenient lasers, they are not in true thermodynamic equilibrium state \cite{10,11}.

Non-equilibrium, or, thermodynamically quasi-equilibrium features of exciton polariton condensates play certain role in various manifestations of their collective (many body) quantum states such as superfluidity \cite{12,13}, quantized vortices \cite{14,15}, soliton formation \cite{16}, Josephson oscillations and macroscopic self-trapping \cite{17,18}.
Since exciton polariton BEC's in present experiments examined at high enough (up to the room) temperatures the quantum character of such a states should be examined even at equilibrium.

The problem of distinguishability of statistically classical and quantum regimes for exciton polariton condensates behavior is very actual for practical purposes, see e.g. \cite{19}; especially for their possible applications for quantum information science purposes.
Fast switching properties (the typical switching time of a few picoseconds), relatively strong nonlinear response and flexibility to external optical and/or electrical pump, spin degrees of freedom made microcavity polaritons potentially very promising for quantum computation and quantum information processing objectives, see, e.g. \cite{21,22,23}. In particular, we would like to mention here problems of design ultrafast quantum gates \cite{22,23}, creation of relatively long Rabi oscillations \cite{19, 20} and quantum annealing problem \cite{24} where collective (bosonic) character of polariton BEC could be used.
Another promising area of potential applications of quantum exciton polariton states is connected with quantum metrology for which designing of quantum interferometers and gyroscopes possessing high precision measurements with quantum phase states is utilized, cf. \cite{25, 26}.

In the paper we consider generic problem of quantum-classical phase transition with exciton polariton condensates occurring in semiconductor microcavity in the presence of Josephson effect with them, cf. \cite{27}.
In general this area of research is based on dissipative tunneling problem or, tunneling at certain temperatures models, cf. \cite{28,29,30}.
Evidently seminal studies on this topic relay to superconductor devices \cite{31} where macroscopic tunneling plays essential role.
In particular, it worth to mention here Schr\"odinger cat state formation \cite{32} and design of quantum gates for quantum computing with superconductor qubits where macroscopic quantum coherent phase properties are used \cite{33}.
Latterly the problem of quantum tunneling and quantum-classical phase transition problem have been applied to other two-level (spin) systems, like atomic BEC's \cite{34}, and small ferromagnetic particles, \cite{35}.
Strictly speaking main discussion is focused on investigation of quantum-classical escape rate transition that takes place in the presence of potential barrier at finite temperatures.

In the paper we propose \textit{extended model} for exciton polariton Josephson junctions when energy of polariton-polariton scattering contributes into the tunneling parameter, cf. \cite{36,37,38} and references therein.
In particularly, we pay our attention to temperature dependent critical phenomena occurring in the presence of macroscopic tunneling.
We aimed at elucidation of necessary criteria to obtain quantum phase regimes with trapped exciton polaritons suitable for creation of exciton polariton qubit gates.

@@

Hamiltonian (\ref{eq:hamiltonian}) represents extended model for internal Josephson junction effect with exciton polariton BEC trapped in double well potential.
It is implies existence of quantum tunneling of the particles between wells with effective (nonlinear) tunneling rate
%
\begin{equation}
G_{eff} = \dfrac{1}{2}(G+\Gamma N - C\hat{\psi}_1^\dag\hat{\psi}_2).
\label{eq:tunneling_rate}
\end{equation}
%
The contribution from relevant coefficients on can be estimated from variational ansatz
%
\begin{equation}
\Phi_{\pm} = A_{\pm} \Big[ \exp \Big( -\dfrac{(x - x_0)^2}{2 a^2} \Big) \pm \exp \Big( -\dfrac{(x + x_0)^2}{2 a^2} \Big) \Big]
\label{eq:two_modes_eq}
\end{equation}
%
where $a$ is width of condensate wavefunctions at each wells (we suppose them equal to each other), defined from normalization conditions.
In Fig. \ref{pic:gamma_pm_vs_g} the behavior of normalized coefficients $\dfrac{\gamma_{ij}}{g}$ $i,j \in \{+,-\}$ as a function of half of intra-well distance $x_0$ is shown.
It is important that at differences between them rapidly vanishes with increasing of distance $x_0$.
We can suppose $\Gamma = C = 0$ for $x_0 >> a$.
This limit exactly corresponds to familiar problem of two weakly linked condensates at zero temperature {\red see e.g. \cite{39} and cf. }.
%
\begin{figure}[ht]
\center{\includegraphics[width=0.65\linewidth]{pic/ypm_x0.png}}
\caption{Dependence of $\gamma_{ij} / g$ versus $x_0 / a$ \label{pic:gamma_pm_vs_g}}
\end{figure}
%

@@

Hamiltonian (\ref{eq:hamiltinian_s}) corresponds to Schrodinger equation
%
\begin{equation}
\begin{array}{l}
(\alpha - \beta \sin^2 \phi) \dfrac{d^2 \Phi}{d \phi^2} + (\beta(s - \dfrac{1}{2})\sin{2\phi}-B\sin{\phi})) \dfrac{d \Phi}{d \phi} \\
+ (E + BS \cos \phi + \beta S^2 \sin^2 \phi + \beta S \cos^2 \phi) = 0.
\end{array}
\label{eq:schrodinger}
\end{equation}
%
where $\Phi \equiv \Phi(\phi)$ is unknown $2\pi$-periodic wavefunction that characterizes quantum phase properties, cf. \cite{40}.
Noticing that Eqs. (\ref{eq:hamiltinian_s}), (\ref{eq:schrodinger}) can be obtained by applying so-called phase-state representation, {\red cf. [  ]}.
%
\begin{equation}
\ket{\psi} = \dfrac{1}{2 \pi} \int_{-\infty}^{+\infty} d \phi \Phi(\phi) \ket{\phi}
\end{equation}
%
where
%
\begin{equation}
\ket{\phi} = \dfrac{1}{\sqrt{N!}} \Big[ e^{i\phi/2} \psi_1^\dagger + e^{-i \phi/2} \psi_2^\dag \Big]^N \ket{0}_1\ket{0}
\end{equation}
%
is macroscopic coherent spin-state (macroscopic qubit state) that plays important role in current quantum information protocols operating with $N$-particle condensates, cf. \cite{41}.
It is possible to eliminate the term with first derivative in Eq.(\ref{eq:schrodinger}) applying the transformation
%
\begin{equation}
\Phi(\phi) = \Psi(z) \exp \Big( s \log \dn{z} - \dfrac{\Lambda s}{2 \sqrt{\lambda (1 - \lambda)}} \arctan \Big( \sqrt{\dfrac{\lambda}{1 - \lambda}} \cn{z} \Big) \Big)
\label{eq:transformation_of_phi}
\end{equation}
%
where
%
\begin{equation}
z = \int \limits_0^\phi \dfrac{d \xi}{\sqrt{1 - \lambda \sin^2 \xi}} = F(\phi, \sqrt{\lambda})
\end{equation}
is new phase variable; $F(\phi, \sqrt{\lambda})$ is incomplete elliptic integral of the first kind.
In (\ref{eq:transformation_of_phi}) we have introduced dimension-less vital parameters $\Lambda = \dfrac{G}{\alpha s}$, $\lambda = \dfrac{\beta}{\alpha}$ completely characterizing our model.

@@

Analysis of the quantum phase for exciton polariton junctions we examine in three domains of  governed parameter $\Lambda$, cf. \cite{40}.
The dependences of trapping potential $V_0(x)$ are shown in Fig. \ref{phase_potential}.
The period of the functions is determined as $4K(\lambda)$.
%
\begin{figure}[ht]
\begin{minipage}[htbp]{0.32\linewidth}	{\includegraphics[width=1\linewidth]{pic/potential_L=100.png} \\ a}
\end{minipage}
\begin{minipage}[htbp]{0.32\linewidth}	{\includegraphics[width=1\linewidth]{pic/potential_L=05.png} \\ b}
\end{minipage}
\begin{minipage}[htbp]{0.32\linewidth}	{\includegraphics[width=1\linewidth]{pic/potential_L=001.png} \\ c}
\end{minipage}
\caption{
Effective potential $V_0(z)$ for (a) $\Lambda = 100$, (b) $\Lambda = 0.5$ , and (c) $\Lambda = 0.01$ as a function of  elliptic integral phase coordinate $z$.
Points $A, B$ and $C$ in (b) corresponds to minima of the potential occurring with periodicity $4K(\lambda)$. \label{phase_potential}}
\end{figure}
%

@@

Another limit that could be studied within Fock regime is relevant to condition $\lambda \ll \Lambda \ll 1$.
In this case from (\ref{eq:potential}) we have
% 
\begin{equation}
V_0(z) \simeq -\Lambda \cos{z}
\label{eq:fock2_potential}
\end{equation}
%
Equation (\ref{eq:fock2_potential}) deals with negligible $\lambda$; it is reproduces results for convenient quantum phase model \cite{40}.

>>>>

For last decade investigations of exciton polariton Bose-Einstein condensates (BEC) in various type of semiconductor microstructures represent one of mainstreams of current photonic and semiconductor technologies, see e.g. \cite{Sanvitto,Guillet}.
The microcavity exciton polaritons are bosonic quasiparticles representing  admixtures of the quantized cavity mode and quantum wells excitons.
Practically, semiconductor microcavities are promising for various optoelectronic applications where quantum  matter-field interface plays crucial role.
In particular, demonstrations of polariton laser with electrical pump \cite{Bhattacharya,Schneider}, amplifiers \cite{Niemietz}, switches \cite{Amo_2010}, transistors \cite{Ballarini}, polariton circuits and optical logic elements \cite{Sturm,Liew}.
Although exciton polariton condensates exhibits Bose-Einstein statistics under the phase transition condition and macroscopic occupation of the ground state at certain pumping rate that is less than for convenient lasers, they are not in true thermodynamic equilibrium state \cite{Byrnes_2014,Sun}.

Non-equilibrium, or, thermodynamically quasi-equilibrium features of exciton polariton condensates play certain role in various manifestations of their collective (many body) quantum states such as superfluidity \cite{Carusotto_2013,Amo_2009}, quantized vortices \cite{Lagoudakis_2008,Lagoudakis_2009}, soliton formation \cite{Sich}, Josephson oscillations and macroscopic self-trapping \cite{Abbarchi,Lagoudakis_2010}.
Since exciton polariton BEC's in present experiments examined at high enough (up to the room) temperatures the quantum character of such a states should be examined even at equilibrium.

The problem of distinguishability of statistically classical and quantum regimes for exciton polariton condensates behavior is very actual for practical purposes, see e.g. \cite{Dominici}; especially for their possible applications for quantum information science purposes.
Fast switching properties (the typical switching time of a few picoseconds), relatively strong nonlinear response and flexibility to external optical and/or electrical pump, spin degrees of freedom made microcavity polaritons potentially very promising for quantum computation and quantum information processing objectives, see, e.g. \cite{Demirchyan,Pagel,Kyriienko,Solnyshkov_2015}. In particular, we would like to mention here problems of design ultrafast quantum gates \cite{Kyriienko,Solnyshkov_2015}, creation of relatively long Rabi oscillations \cite{Dominici, Demirchyan} and quantum annealing problem \cite{Yan} where collective (bosonic) character of polariton BEC could be used.
Another promising area of potential applications of quantum exciton polariton states is connected with quantum metrology for which designing of quantum interferometers and gyroscopes possessing high precision measurements with quantum phase states is utilized, cf. \cite{Gulevich, Pezze}.

In the paper we consider generic problem of quantum-classical phase transition with exciton polariton condensates occurring in semiconductor microcavity in the presence of Josephson effect with them, cf. \cite{Chudnovsky_1997}.
In general this area of research is based on dissipative tunneling problem or, tunneling at certain temperatures models, cf. \cite{Caldeira,Larkin,Riseborough}.
Evidently seminal studies on this topic relay to superconductor devices \cite{Ankerhold} where macroscopic tunneling plays essential role.
In particular, it worth to mention here Schr\"odinger cat state formation \cite{Leggett} and design of quantum gates for quantum computing with superconductor qubits where macroscopic quantum coherent phase properties are used \cite{Makhlin}.
Latterly the problem of quantum tunneling and quantum-classical phase transition problem have been applied to other two-level (spin) systems, like atomic BEC's \cite{Zhang}, and small ferromagnetic particles, \cite{Owerre}.
Strictly speaking main discussion is focused on investigation of quantum-classical escape rate transition that takes place in the presence of potential barrier at finite temperatures.

In the paper we propose \textit{extended model} for exciton polariton Josephson junctions when energy of polariton-polariton scattering contributes into the tunneling parameter, cf. \cite{Borgh,Solnyshkov_2008,Sarchi} and references therein.
In particularly, we pay our attention to temperature dependent critical phenomena occurring in the presence of macroscopic tunneling.
We aimed at elucidation of necessary criteria to obtain quantum phase regimes with trapped exciton polaritons suitable for creation of exciton polariton qubit gates.

@@

Hamiltonian (\ref{eq:hamiltonian}) represents extended model for internal Josephson junction effect with exciton polariton BEC trapped in double well potential.
It is implies existence of quantum tunneling of the particles between wells with effective (nonlinear) tunneling rate
%
\begin{equation}
G_{eff} = \dfrac{1}{2}(G+\Gamma N - C\hat{\psi}_1^\dag\hat{\psi}_2).
\label{eq:tunneling_rate}
\end{equation}
%
The contribution from relevant coefficients on can be estimated from variational ansatz
%
\begin{equation}
\Phi_{\pm} = A_{\pm} \Big[ \exp \Big( -\dfrac{(x - x_0)^2}{2 a^2} \Big) \pm \exp \Big( -\dfrac{(x + x_0)^2}{2 a^2} \Big) \Big]
\label{eq:two_modes_eq}
\end{equation}
%
where $a$ is width of condensate wavefunctions at each wells (we suppose them equal to each other), defined from normalization conditions.
In Fig. \ref{pic:gamma_pm_vs_g} the behavior of normalized coefficients $\dfrac{\gamma_{ij}}{g}$ $i,j \in \{+,-\}$ as a function of half of intra-well distance $x_0$ is shown.
It is important that at differences between them rapidly vanishes with increasing of distance $x_0$.
We can suppose $\Gamma = C = 0$ for $x_0 >> a$.
This limit exactly corresponds to familiar problem of two weakly linked condensates at zero temperature {\red see e.g. \cite{Raghavan} and cf. }.
%
\begin{figure}[ht]
\center{\includegraphics[width=0.65\linewidth]{pic/ypm_x0.png}}
\caption{Dependence of $\gamma_{ij} / g$ versus $x_0 / a$ \label{pic:gamma_pm_vs_g}}
\end{figure}
%

@@

Hamiltonian (\ref{eq:hamiltinian_s}) corresponds to Schrodinger equation
%
\begin{equation}
\begin{array}{l}
(\alpha - \beta \sin^2 \phi) \dfrac{d^2 \Phi}{d \phi^2} + (\beta(s - \dfrac{1}{2})\sin{2\phi}-B\sin{\phi})) \dfrac{d \Phi}{d \phi} \\
+ (E + BS \cos \phi + \beta S^2 \sin^2 \phi + \beta S \cos^2 \phi) = 0.
\end{array}
\label{eq:schrodinger}
\end{equation}
%
where $\Phi \equiv \Phi(\phi)$ is unknown $2\pi$-periodic wavefunction that characterizes quantum phase properties, cf. \cite{Anglin}.
Noticing that Eqs. (\ref{eq:hamiltinian_s}), (\ref{eq:schrodinger}) can be obtained by applying so-called phase-state representation, {\red cf. [  ]}.
%
\begin{equation}
\ket{\psi} = \dfrac{1}{2 \pi} \int_{-\infty}^{+\infty} d \phi \Phi(\phi) \ket{\phi}
\end{equation}
%
where
%
\begin{equation}
\ket{\phi} = \dfrac{1}{\sqrt{N!}} \Big[ e^{i\phi/2} \psi_1^\dagger + e^{-i \phi/2} \psi_2^\dag \Big]^N \ket{0}_1\ket{0}
\end{equation}
%
is macroscopic coherent spin-state (macroscopic qubit state) that plays important role in current quantum information protocols operating with $N$-particle condensates, cf. \cite{Byrnes_2012}.
It is possible to eliminate the term with first derivative in Eq.(\ref{eq:schrodinger}) applying the transformation
%
\begin{equation}
\Phi(\phi) = \Psi(z) \exp \Big( s \log \dn{z} - \dfrac{\Lambda s}{2 \sqrt{\lambda (1 - \lambda)}} \arctan \Big( \sqrt{\dfrac{\lambda}{1 - \lambda}} \cn{z} \Big) \Big)
\label{eq:transformation_of_phi}
\end{equation}
%
where
%
\begin{equation}
z = \int \limits_0^\phi \dfrac{d \xi}{\sqrt{1 - \lambda \sin^2 \xi}} = F(\phi, \sqrt{\lambda})
\end{equation}
is new phase variable; $F(\phi, \sqrt{\lambda})$ is incomplete elliptic integral of the first kind.
In (\ref{eq:transformation_of_phi}) we have introduced dimension-less vital parameters $\Lambda = \dfrac{G}{\alpha s}$, $\lambda = \dfrac{\beta}{\alpha}$ completely characterizing our model.

@@

Analysis of the quantum phase for exciton polariton junctions we examine in three domains of  governed parameter $\Lambda$, cf. \cite{Anglin}.
The dependences of trapping potential $V_0(x)$ are shown in Fig. \ref{phase_potential}.
The period of the functions is determined as $4K(\lambda)$.
%
\begin{figure}[ht]
\begin{minipage}[htbp]{0.32\linewidth}	{\includegraphics[width=1\linewidth]{pic/potential_L=100.png} \\ a}
\end{minipage}
\begin{minipage}[htbp]{0.32\linewidth}	{\includegraphics[width=1\linewidth]{pic/potential_L=05.png} \\ b}
\end{minipage}
\begin{minipage}[htbp]{0.32\linewidth}	{\includegraphics[width=1\linewidth]{pic/potential_L=001.png} \\ c}
\end{minipage}
\caption{
Effective potential $V_0(z)$ for (a) $\Lambda = 100$, (b) $\Lambda = 0.5$ , and (c) $\Lambda = 0.01$ as a function of  elliptic integral phase coordinate $z$.
Points $A, B$ and $C$ in (b) corresponds to minima of the potential occurring with periodicity $4K(\lambda)$. \label{phase_potential}}
\end{figure}
%

@@

Another limit that could be studied within Fock regime is relevant to condition $\lambda \ll \Lambda \ll 1$.
In this case from (\ref{eq:potential}) we have
% 
\begin{equation}
V_0(z) \simeq -\Lambda \cos{z}
\label{eq:fock2_potential}
\end{equation}
%
Equation (\ref{eq:fock2_potential}) deals with negligible $\lambda$; it is reproduces results for convenient quantum phase model \cite{Anglin}.
